---
title: "Probability, Entropy and Inference"
date: 03/07/2005
date-format: "DD MMMM YYYY"
date-modified: last-modified
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
#| label: imports-and-functions-r
#| code-summary: "Imports and Functions - R"
#| output: false

library(tidyverse)
library(reticulate)
library(ggpubr)
library(ggthemes)
library(kableExtra)
```


---

### Exercise 2.2

::: {.callout-note}
Are the random variables X and Y in the joint ensemble of figure 2.2 independent?
:::

Yes, because the marginal probabilities of monograms marginalized over the joint ensemble on probabilities for all possible bigrams $P(x,y)$ equal exactly the probability of monograms themselves $P(x) = P(y)$.


### Exercise 2.4

::: {.callout-note}
An urn contains $K$ balls, of which $B$ are black and $W = K − B$ are white. Fred draws a ball at random from the urn and replaces it, $N$ times.
a) What is the probability distribution of the number of times a black ball is drawn, $nB$?
a) What is the expectation of $nB$? What is the variance of $nB$? What is the standard deviation of $nB$? Give numerical answers for the cases $N = 5$ and $N = 400$, when $B = 2$ and $K = 10$.
:::



### Exercise 2.4

::: {.callout-note}
An urn contains $K$ balls, of which $B$ are black and $W = K−B$ are white. We define the fraction $fB≡B/K$.
Fred draws N times from the urn, exactly as in exercise 2.4, obtaining $nB$ blacks, and computes the quantity
$$
z = \frac{(n_B-f_{B}N)^2}{Nf_B(1-f_B)}
$$
What is the expectation of $z$? In the case $N = 5$ and $fB = 1/5$,
what is the probability distribution of $z$? What is the probability that $z < 1$?
:::

::: {.callout-tip appearance=default}
Compare z with the quantities computed in the previous exercise.
:::

