---
title: "Chapter 02 - Probability, Entropy and Inference"
date: 03/07/2005
date-format: "DD MMMM YYYY"
date-modified: last-modified
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
#| label: imports-and-functions-r
#| code-summary: "Imports and Functions - R"
#| output: false
#| echo: false

library(tidyverse)
library(reticulate)
library(ggpubr)
library(ggthemes)
library(kableExtra)
library(latex2exp)
```

---

### Exercise 2.2

::: {.callout-note}
Are the random variables X and Y in the joint ensemble of figure 2.2 independent?
:::

Yes, because the marginal probabilities of monograms marginalized over the joint ensemble on probabilities for all possible bigrams $P(x,y)$ equal exactly the probability of monograms themselves $P(x) = P(y)$.


### Exercise 2.4.a

::: {.callout-note}
An urn contains $K$ balls, of which $B$ are black and $W = K − B$ are white. Fred draws a ball at random from the urn and replaces it, $N$ times.

a) What is the probability distribution of the number of times a black ball is drawn, $nB$?
a) What is the expectation of $nB$? What is the variance of $nB$? What is the standard deviation of $nB$? Give numerical answers for the cases $N = 5$ and $N = 400$, when $B = 2$ and $K = 10$.
:::

Since this is a sampling with replacement, each draw is independent and can be considered a Bernoulli trial with outcome black or white.

The distribution of aggregate outcomes of a fixed number of Bernoulli trials (i.e. how many of the drawn balls are expected to be black or white) with a constant rate of success follows a binomial distribution with _pdf_:

$$
f(n,k,p) = {n \choose k}p^{k}(1-p)^{n-k}
$$

Where $n$ is the number of trials, $k$ is the number of successes (black balls drawn) and $p$ is the success rate $B/K$.

#### Expectation and Mean

The mean is simply $np$ as each draw we have probability $p$ of drawing a black ball.
The mode (expectation) is the mean rounded to the nearest integer.

#### Variance

This is a bit more complex and requires a few tricks. Lets start with the general definition of variance

$$
\mathbb{V} = \mathbb{E}[X^2] - \mathbb{E}[X]
$$ {#eq-general-variance}

As we introduced above, $\mathbb{E}[X] = np$.


Solving $\mathbb{E}[X^2]$ directly will be too complex so we resort to factorizing it down to a combination of lower order factorial moments. Big words to say: _something that is equivalent but lower power_.

$$
\mathbb{E}[X^2] = \mathbb{E}[X(X - 1)] + \mathbb{E}[X]
$$

This is nice because $\mathbb{E}[X]$ is again simply the mean $np$.

The first term is akin to having two trial $X$ and $X-1$ which means it can be simplified to

$$
\mathbb{E}[X(X - 1)] = n(n-1)p^2
$$

Substituting all the terms to @eq-general-variance we get

$$
\begin{split}
\mathbb{V} &= \mathbb{E}[X^2] - \mathbb{E}[X] \\
&= n(n-1)p^2 + np - n^2p^2 \\
&= \cancel{n^2p^2} - np^2 + np - \cancel{n^2p^2} \\
&= np(1-p)
\end{split}
$$

And the standard deviation is $\sigma = \sqrt{\mathbb{V}} = \sqrt{np(1-p)}$.

#### Numerical answers

- $N = 5$ the expectation is `r round(5*(2/10))` and the standard deviation is `r 5*(2/10)*(1 - (2/10))`
- $N = 400$ the expectation is `r round(400*(2/10))` and the standard deviation is `r 400*(2/10)*(1 - (2/10))`

```{r}
#| fig-cap: Binomial distributions for drawing 10 and 400 balls with replacement
#| label: fig-2.4-binomials-sampling-with-replacement
#| fig-subcap:
#|   - "For 5 draws"
#|   - "For 400 draws"
#| layout-ncol: 1

n_black <- 2
n_balls <- 10
p_black <- n_black / n_balls

for (n_draws in c(5, 400)) {
  black_balls <- c(0:n_draws)
  p <- data.frame(black_balls = black_balls) %>%
    mutate(
      prob_success = dbinom(black_balls, n_draws, p_black)
    ) %>%
    ggplot(aes(black_balls, prob_success)) +
    geom_line(color = "black", alpha = .3) +
    geom_point(color = "red", size = 1) +
    ggthemes::theme_few()
  print(p)
}
```


### Exercise 2.4.b

::: {.callout-note}
An urn contains $K$ balls, of which $B$ are black and $W = K−B$ are white. We define the fraction $f{B}\equiv{B}/K$.
Fred draws N times from the urn, exactly as in exercise 2.4, obtaining $nB$ blacks, and computes the quantity

$$
z = \frac{(n_B-f_{B}N)^2}{Nf_B(1-f_B)}
$$

What is the expectation of $z$? In the case $N = 5$ and $fB = 1/5$,
what is the probability distribution of $z$? What is the probability that $z < 1$?
:::

::: {.callout-tip appearance=default}
Compare z with the quantities computed in the previous exercise.
:::

To compute the expectation of $z$, we can substitute the expectation of $n_B$ which is the expectation for the binomial we described above ($np$) to the equation

$$
n_B = f_{B}N
$$

From which follows the simplification

$$
z = \frac{(\cancel{f_{B}N}-\cancel{f_{B}N})^2}{Nf_B(1-f_B)} = 0
$$

At the point of maximum likelihood for the binomial, z is 0. Let's plot the distribution (the parameters are the same as the last exercise when drawing 5 balls, as the probability of drawing a black ball is still 0.2).

```{r}
#| fig-cap: $z$ distribution for drawing 5 balls with replacement. Dashed line represents $z = 1$.
#| label: fig-2.4-z-sampling-with-replacement

fz <- function(n_blacks, n_draws, p_black) {
  (n_blacks - p_black * n_draws)**2 /
    (n_draws * p_black * (1 - p_black))
}

n_draws <- 5
p_black <- n_black / n_balls
black_balls <- c(0:n_draws)

binomial_vs_z <- data.frame(black_balls = black_balls) %>%
  mutate(z = fz(black_balls, n_draws, p_black))

binomial_vs_z %>%
  ggplot(aes(black_balls, z)) +
  geom_line(alpha = .7) +
  geom_point(size = 1, color = "red") +
  geom_hline(
    yintercept = 1.0,
    linetype = "dashed",
    color = "grey",
    alpha = 0.5
  ) +
  ggthemes::scale_color_few() +
  ggthemes::theme_few()
```

To find out the probability of $z < 1$, let's rework the equation for $z$ and substitute the constant terms

$$
\begin{split}
\frac{(n_B-f_{B}N)^2}{Nf_B(1-f_B)} &< 1 \\
(n_B-1)^2 &< 1(1-0.2) \\
(n_B-1)^2 &< 0.8 \\
\end{split}
$$

The function of $z$ is less than $1$ only when $0.2$ as, as $n_B$ can only assume integer values between $1$ and $5$, and the left side can only be less than $1$ when $n_B-1 = 0$, $z<1$ when $n_B = 1$, which is also the expected number of black balls.

This is also evident when plotting the $z$ function in @fig-2.4-z-sampling-with-replacement.

### Exercise 2.8

::: {.callout-note appearance=default}

The problem in example 2.7 that we are trying to solve is the following:

Bill tosses a bent coin $N$ times, obtaining a sequence of heads
and tails. We assume that the coin has a probability $f_H$ of coming up
heads; we do not know $f_H$. If $n_H$ heads have occurred in $N$ tosses, what
is the probability distribution of $f_H$? (For example, $N$ might be 10, and
$n_H$ might be 3; or, after a lot more tossing, we might have $N = 300$ and
$n_H = 29$.) What is the probability that the $N+1^{th}$ outcome will be a
head, given $n_H$ heads in $N$ tosses?

:::

::: {.callout-note}

Assuming a uniform prior on $f_H$, $P(f_H) = 1$, solve the problem
posed in example 2.7 (p.30). Sketch the posterior distribution of $f_H$ and
compute the probability that the $N+1^{th}$ outcome will be a head, for

a) $N$ = 3 and $n_H$ = 0;
b) $N$ = 3 and $n_H$ = 2;
c) $N$ = 10 and $n_H$ = 3;
d) $N$ = 300 and $n_H$ = 29.

:::

::: {.callout-tip appearance=default}

You will find the beta integral useful:

$$
\int_0^1 \mathrm{d}p_a\, p_a^{F_a}(1 - p_a)^{F_b} = \frac{\Gamma(F_a + 1)\Gamma(F_b + 1)}{\Gamma(F_a + F_b + 2)} = \frac{F_a!F_b!}{(F_a + F_b + 1)!}
$$

:::

Let's start by formalizing the problem in Bayesian terms. The probability of the next coin being heads it's equal to the expected rate of heads for the coin itself. We do not know the rate but we can estimate the probability function of the rate from the coin flips that already happened.

$$
p(f_H|n_H,N) \propto P(n_H|f_H,N) P(f_H)
$$

$N$ is a known variable so it stays to the left of the proposition.

Given that the prior for $f_H$ is uniform, we can ignore the prior and the equation simplifies to

$$
p(f_H|n_H,N) \propto P(n_H|f_H,N)
$$

$P(n_H|f_H,N)$ can be computed using the binomial distribution:

> The probability of "heads" given a rate of heads for the coin (its "fairness") and the number of times we flip the coin

$$
P(n_H|f_H,N) = {N \choose n_H}{f_H}^{n_H}(1-f_H)^{N-n_H}
$$
Which we can substitute above and drop $N \choose n_H$, as it is a constant term across all values of $f_H$.
$$
P(f_H|n_H,N) \propto {f_H}^{n_H}(1-f_H)^{N-n_H}
$$ {#eq-bayes-rule-bent-coin}

What we have now is simply the kernel of the binomial distribution: the beta distribution. Luckily there are easy ways to solve for $f_H$ (one is the beta integral above). The integral is

$$
P(N+1=H) = P(f_H|n_H,N) = \int_0^1{d{f_H}}{f_H}^{n_H}(1-f_H)^{N-n_H}
$$


::: {.callout-note appearance=default}

The _kernel_ of a distribution is the *part of a probability density that contains all the dependence on the variable*.

:::

Let's plot the beta distribution for each of the above draws and compute the probability of heads for the $N+1$ draw.


```{r}
#| fig-cap: Sketch of the posterior distribution of $f_H$ given a fixed number of draws and a fixed number of heads.
#| label: fig-2.8-new-draw-bent-coin

beta <- function(draws, heads, fh) {
  fh**heads * (1 - fh)**(draws - heads)
}

df <- data.frame(
  question = c(
    r"($N$ = 3 and $n_H$ = 0)",
    r"($N$ = 3 and $n_H$ = 2)",
    r"($N$ = 10 and $n_H$ = 3)",
    r"($N$ = 300 and $n_H$ = 29)"
  ),
  draws = c(3, 3, 10, 300),
  heads = c(0, 2, 3, 29)
) %>%
  expand_grid(fh = seq(0.0, 1.0, 0.01)) %>%
  mutate(lfh = beta(draws, heads, fh)) %>%
  group_by(question) %>%
  mutate(sum_lfh = sum(lfh)) %>%
  mutate(pfh = lfh / sum_lfh)

p <- df %>%
  ggplot(aes(fh, pfh, color = question)) +
  geom_line(alpha = .3) +
  geom_point(size = 0.5) +
  ggthemes::scale_color_few(
    name = "Coin Flips",
    labels = TeX(df$question)
  ) +
  ggthemes::theme_few() +
  theme(
    legend.text = element_text(hjust = 0),
    legend.position = "bottom"
  ) +
  labs(
    x = TeX(r"($f_H$)"),
    y = TeX(r"($P(f_H|N,n_H)$)"),
  )

suppressMessages(suppressWarnings(print(p)))
```

We sketched the posteriors in @fig-2.8-new-draw-bent-coin Now let's compute the probability of heads for the next draw as given by the beta integral in each case (solved between 0 and 1).

```{r}
#| label: tbl-2.8-probability-next-head-bent-coin
#| tbl-cap: Probability of heads for the next draw given the previous draws.

beta_integral <- function(draws, heads) {
  (heads + 1) / (draws + 2)
}

data.frame(
  Draws = c(3, 3, 10, 300),
  Heads = c(0, 2, 3, 29)
) %>%
  mutate(`$P(N+1=Heads)$` = round(beta_integral(Draws, Heads), 3)) %>%
  knitr::kable(align = "ccc")
```

@tbl-2.8-probability-next-head-bent-coin shows the probabilities of obtaining heads on the next draw given different sets of coin flips.

I have to say I made it a lot more simplistic than it is. The proper way to compute the values in @tbl-2.8-probability-next-head-bent-coin would have been to compute the $P(B)$ in the full Bayesian equation at @eq-bayes-rule-bent-coin instead of just ignoring it and use proportionality.

The full derivation simplifies to $(n_H+1)/(N+2)$ for uniform priors (which I used directly in @tbl-2.8-probability-next-head-bent-coin.

### Exercise 2.10

::: {.callout-note}

Urn A contains three balls: one black, and two white; urn B contains three balls: two black, and one white. One of the urns is selected at random and one ball is drawn. The ball is black. What is the probability that the selected urn is urn A?

:::

### Exercise 2.11

::: {.callout-note}

Urn A contains five balls: one black, two white, one green and one pink; urn B contains five hundred balls: two hundred black, one hundred white, 50 yellow, 40 cyan, 30 sienna, 25 green, 25 silver, 20 gold, and 10 purple. In other words: One fifth of A’s balls are black, two-fifths of B’s are black. 

One of the urns is selected at random and one ball is drawn. The ball is black. What is the probability that the urn is urn A?

:::
